---
title: "Rust style enums"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Rust style enums}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

# The problem

Strings are R's enums. `svd_solver = "randomised"`, a `match.arg()`, a
`switch()` somewhere downstream. Fine while the variants carry nothing. It falls
apart the moment one of them does. Take an optimiser argument: `"sgd"` wants a
`momentum`, `"adam"` wants `beta1` and `beta2`, `"lbfgs"` wants a `history`
size. Put them all flat in one params list and pass `momentum` with `"adam"`:
it gets silently ignored. The roxygen ends up saying "only used for `sgd`" and
"ignored for `lbfgs`", which is documentation doing the job of a type.

Rust solves this with data-carrying enums: each variant owns its fields, a
`match` has to cover every variant, and `impl` blocks hang behaviour off the
type. `devforge` generates the closest R equivalent. Same deal as the params
wrappers: a spec goes in, plain `checkmate` code comes out, and no package
gains a runtime dependency on `devforge`.

```{r}
#| label: setup
library(devforge)
```

# Declaring an enum

A `param_enum()` holds the variants, each a `p_variant()` with its own doc and
fields. The fields are the same `p_int()` and friends the params specs use.

```{r}
#| label: solver
pca_solver <- param_enum(
  name = "pca_solver",
  title = "PCA solver",
  variants = list(
    covariance = p_variant("Eigendecomposition of the cross-product."),
    randomised = p_variant(
      "Randomised SVD. Approximate, the trailing components the least so.",
      oversample = p_int(10L, range = "[0,)", doc = "Extra columns sampled."),
      n_iter = p_int(2L, range = "[0,)", doc = "Power iterations.")
    ),
    exact = p_variant("Lanczos on sparse input, a full SVD on dense.")
  )
)
```

Every variant gets a constructor, validated like any `params_*()` function:

```{r}
#| label: ctor
cat(devforge:::emit_variant_ctor(pca_solver, "randomised"), sep = "\n")
```

A value is a plain list. `variant` is an element, not an attribute, because
`c()` and `modifyList()` drop attributes, and because a compiled backend reads
a list element without any ceremony: `list.get("variant")` in Rust, say. The
class vector is there for S3 dispatch, see below.

To play with the generated code, render it and evaluate it into an environment.
In a package, `forge_params()` writes it to `R/enums-generated.R` instead.

```{r}
#| label: eval
env <- new.env()
rendered <- devforge:::render_specs(list(pca_solver = pca_solver))
eval(parse(text = rendered$enums), envir = env)

str(env$pca_solver_randomised(n_iter = 4L))
```

# Coercion, checking and matching

`as_pca_solver()` takes a variant name and builds it with its defaults, so
callers keep typing strings. Hand it a list and it gets checked and its class
rebuilt, which is what you want after a round trip through Rust or `saveRDS()`.

```{r}
#| label: as
class(env$as_pca_solver("exact"))
class(env$as_pca_solver(unclass(env$pca_solver_randomised())))
```

The checker is strict about names. A field that belongs to another variant is
an error, which is the entire point:

```{r}
#| label: check
x <- env$pca_solver_exact()
x$n_iter <- 4L
env$checkPcaSolver(x)
```

`match_pca_solver()` takes one function per variant. Each gets the whole value,
so fields are read by name. Reorder the fields in the spec and nothing silently
swaps.

```{r}
#| label: match
n_components <- \(solver, k) {
  env$match_pca_solver(
    solver,
    covariance = \(v) k,
    randomised = \(v) k + v$oversample,
    exact = \(v) k
  )
}
n_components("randomised", 30L)
```

Leave an arm out and it errors on every call, not just when the missing
variant turns up. That is as close to Rust's compile-time check as R gets:

```{r}
#| label: match-missing
#| error: true
env$match_pca_solver("covariance", covariance = \(v) 1)
```

`.default` is the `_ =>` arm, for when you mean it:

```{r}
#| label: match-default
env$match_pca_solver(
  "exact",
  randomised = \(v) "approx",
  .default = \(v) "exact"
)
```

Why not [matchr](https://cran.r-project.org/package=matchr)? It shares the
representation, but its `Match()` only errors when the value in hand hits no
arm, destructures by position, does no validation, and every variant name used
in an arm trips `R CMD check` with "no visible global function definition".

# Impl blocks

Declare methods on the enum and `devforge` emits one S3 generic per method:

```{r}
#| label: methods
pca_solver <- param_enum(
  name = "pca_solver",
  title = "PCA solver",
  variants = pca_solver$variants,
  methods = list(is_exact = "Is the solver exact")
)
cat(devforge:::emit_enum_generics(list(pca_solver)), sep = "\n")
```

The bodies you write by hand under `R/`, as ordinary S3 methods. Per variant
class, or once for the enum class as the fallback arm:

```r
#' @export
is_exact.PcaSolver <- function(self, ...) TRUE

#' @export
is_exact.PcaSolver_Randomised <- function(self, ...) FALSE
```

They stay in `R/` on purpose: linted, covered, debuggable. What `devforge` adds
is the exhaustiveness check. `forge_params()` and `params_up_to_date()` parse
`R/` and error when a variant has neither its own method nor the fallback:

```
Error: Enum methods without an implementation. Add a method per variant or one for the enum class:
  is_exact() on PcaSolver: covariance, exact
```

Add a variant to the spec, forget the method, and the drift CI goes red. Pretty
much what `rustc` would tell you, minus the compile.

# Enums in params

A `param_spec()` field takes an enum with `p_enum()`. The formal defaults to the
variant name, the constructor coerces with `as_pca_solver()`, and the returned
list carries the full variant:

```{r}
#| label: params
spec_pca <- param_spec(
  name = "pca",
  title = "Wrapper function for the PCA parameters",
  fields = list(
    no_pcs = p_int(30L, range = "[1,)", doc = "Number of PCs."),
    svd_solver = p_enum("pca_solver", "covariance", doc = "Which solver.")
  )
)
cat(devforge:::emit_ctor(spec_pca), sep = "\n")
```

The params checker hands the field to `checkPcaSolver()`, so a stray `n_iter`
on the `"exact"` solver fails there as well.

```{r}
#| label: params-eval
rendered <- devforge:::render_specs(list(
  pca = spec_pca,
  pca_solver = pca_solver
))
env <- new.env()
eval(parse(text = unlist(rendered)), envir = env)

str(env$params_pca(svd_solver = "randomised")$svd_solver)
```

One restriction: `param_defaults()` blocks cannot carry an enum. They get merged
with `modifyList()`, which recurses into the tagged list and would happily glue
the `adam` fields onto an `sgd` variant. Shared defaults blocks stay flat for
now.

# Beyond parameters

Nothing about the enums is params specific. The params side is one field type
on top. A variant field can be anything a spec field can be, bar `p_merge()`:

- no default, and the constructor requires it
- `p_free()`, for a matrix, a `data.table` or anything else a pattern cannot
  express; the checker only demands it is there
- `p_enum()`, for a nested enum, checked with that enum's own checker

That makes result types straightforward. A PCA run either finished, carrying
its scores and the solver that produced them, or got skipped for a reason:

```{r}
#| label: outcome
pca_outcome <- param_enum(
  name = "pca_outcome",
  title = "PCA outcome",
  variants = list(
    done = p_variant(
      "The decomposition finished.",
      scores = p_free(doc = "Observation x component matrix."),
      solver = p_enum("pca_solver", doc = "The solver that ran.")
    ),
    skipped = p_variant(
      "Nothing to decompose.",
      reason = p_chr(doc = "Why it was skipped.")
    )
  )
)

rendered <- devforge:::render_specs(list(
  pca_outcome = pca_outcome,
  pca_solver = pca_solver
))
env <- new.env()
eval(parse(text = rendered$enums), envir = env)

res <- env$pca_outcome_done(matrix(rnorm(6), 3, 2), solver = "exact")
env$match_pca_outcome(
  res,
  done = \(v) sprintf("%d PCs via %s", ncol(v$scores), v$solver$variant),
  skipped = \(v) v$reason
)
```

A variant with required fields cannot be built from its name, and
`as_pca_outcome("done")` tells you which constructor to call instead.

# Running it

Enums live in `inst/params/*.R` next to the specs, and the same
`forge_params()` call picks them up. They go into their own file,
`R/enums-generated.R`, with a small self-contained helper at the top, so a
package with enums and no params specs gets just that one file. Delete the
last enum and the next `forge_params()` removes the file; `params_up_to_date()`
flags it as stale until then.

# What is next

The other side of the FFI. Packages that hand their params to Rust through
`extendr` still parse the strings again over there. A generated
`TryFrom<&Robj>` per enum, reading `variant` and its fields, would make the
spec the single source for both languages. And merge semantics for enums
inside defaults blocks, so shared blocks can carry their fields per variant
too. Watch this space.
