---
title: "Generating parameter wrappers"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Generating parameter wrappers}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

# The problem

You like to reduce [parameter clutter for your functions](https://design.tidyverse.org/argument-clutter.html)
according to tidy design principles?  I do. A `params_xxx()` constructor that 
returns a named list, a  `checkXxxParams()` in a `checkmate_extensions.R` file, 
and an `assertXxxParams()` built from it with 
`checkmate::makeAssertionFunction()`. However, this can be a pain in large 
packages with lots and lots of these parameter wrappers.

Writing the same thing a hundred times has consequences: spelling mistakes,
inconsistent naming, documentation drift and staleness, etc. You name it.

`devforge` takes a spec and writes both halves. It is a dev-time tool: the
generated files are committed, they are plain `checkmate` code, and no package
gains a runtime dependency on it.

```{r}
#| label: setup
library(devforge)
```

# A spec

A field carries a default, a type, an optional range, and its roxygen prose.

```{r}
#| label: spec
spec <- param_spec(
  name = "kernel",
  title = "Wrapper function for the diffusion kernel parameters",
  checker = "Kernel",
  label = "kernel parameters",
  fields = list(
    sigma2 = p_dbl(1.0, doc = "Bandwidth parameter."),
    add_diag = p_dbl(1.0, doc = "Regularisation added to the diagonal."),
    a = p_dbl(3.0, doc = "Regularisation parameter for the `pstep` kernel."),
    p = p_int(5L, doc = "Number of steps for the `pstep` kernel.")
  ),
  extra_ctor = quote({
    checkmate::assertTRUE(a >= 2, .var.name = "a >= 2")
    checkmate::assertTRUE(p > 0L, .var.name = "p > 0")
  }),
  extra_check = quote({
    if (x$a < 2) {
      return("Parameter `a` must be >= 2 for the pstep kernel.")
    }
    if (x$p < 1L) {
      return("Parameter `p` must be a positive integer.")
    }
  })
)
```

That is the real `genewalkR::params_kernel()` spec, see [here](https://gregorlueg.github.io/genewalkR/reference/params_kernel.html). 
Note what is *not* in it: the qassert strings, the type annotations in the 
roxygen, the documented defaults, the `must.include` name list, the rule tables, 
and the assert sibling. All of those are derived, which is exactly why they 
cannot drift.

The two `extra_*` slots are the escape hatch. Cross-field rules like `a >= 2`
cannot be expressed as a per-field pattern, so they go in verbatim. The
constructor asserts, the checker returns a string: that is the only difference
between the two, and both come from one place in the spec.

# What comes out

```{r}
#| label: ctor
cat(devforge:::emit_ctor(spec), sep = "\n")
```

The checker is three calls into a small set of shared helpers, which
`devforge` also emits so every package has them:

```{r}
#| label: checker
cat(devforge:::emit_checker(spec), sep = "\n")
```

# Field types

`p_int()`, `p_dbl()`, `p_lgl()`, `p_chr()`, `p_choice()` and `p_free()`. Range
and nullability fold into one qassert pattern, which settles the two competing
spellings on one:

```{r}
#| label: qassert
devforge:::field_qassert(p_int(200L, range = "[1,)"))
devforge:::field_qassert(p_int(NULL, range = "[1,)", null_ok = TRUE))
devforge:::field_qassert(p_int(NULL, len = "+", null_ok = TRUE))
devforge:::field_qassert(p_dbl(1.0, range = "[1, 2]", strict = TRUE))
```

A choice field emits `match.arg()` plus `checkmate::assertChoice()` in the
constructor and an `apply_choice_rules()` entry in the checker, and moves the
default to the front of the set so `match.arg()` picks it:

```{r}
#| label: choice
p_choice("kmknn", c("hnsw", "kmknn", "annoy"))$choices
```

`p_free()` is the hatch for anything a pattern cannot express. It contributes a
formal, a default and its roxygen, and nothing else. Pair it with an
`extra_check`.

# Running it

Specs live in `inst/params/*.R`. They ship, so one package can reference
another's rather than reaching in with `:::`.

```r
devforge::forge_params(".")
```

That writes three files under `R/`, each headed `# Generated by devforge. Do not
edit.`, and runs `air` over them:

- `R/params-generated.R`, the constructors
- `R/checkmate-params-generated.R`, the checkers and their assert siblings
- `R/params-prelude-generated.R`, the three shared helpers

The hand-written checkers that validate something other than a params list stay
where they are. `checkGeneWalkGraphDt()` checks a `data.table`, and that is not
this tool's job.

For CI there is a drift guard that regenerates into a temporary directory and
compares:

```r
devforge::params_up_to_date(".")
```

`TRUE` when everything matches, otherwise the paths that are stale. That catches
both a hand-edit of a generated file and a spec change that was never
regenerated.

# What is next

Three linters, each aimed at something the survey of the five packages turned up
rather than at a hypothetical: roxygen that has drifted away from the formals it
documents, function inputs that never reach a `checkmate` call, and `rs_*`
bindings without an R wrapper. Watch this space.
